---
title: "Cyclistic Bike-Share Analysis"
author: "Ahmed Nasser"
date: "2024-03-03"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## **A summary of the content:**
- Introduction.
-	A clear statement of the business task.
-	A description of all data sources used.
-	Documentation of any cleaning or manipulation of data.
-	A summary of the analysis.
-	Supporting visualizations and key findings.
-	The top three recommendations based on my analysis.


## **Introduction:**
In this project I will analyze the Cyclistic’s data to get insights and answer questions using the six steps of the data life cycle (Ask, Prepare, Process, Analyze, Share, and Act) using R for processing and cleaning, SQL for summarizing and analyzing, Tableau for visualizing the data, and PowerPoint for presenting the insights to answer the business task.

### **1. Ask:** A clear statement of the business task.
#####	We need to get insights to answer: How do annual members and casual riders use Cyclistic bikes differently?
- **But first we need to know:**
  a.	What is the weekly average ride length for casual and members riders?
  b.	Is there a correlation between the seasons in Chicago and the number of rides?
  c.	Is there a correlation between the seasons in Chicago and ride length?
  d.	What is the average number of rides on each day of the week?
  e.	What is the average ride length on each day of the week?
  f.	Which stations do casual riders often visit?
  g.	What is the most used bike_type for casual riders?

### **2. Prepare:** A description of all data sources used.
- I downloaded the previous 12 months data from this site https://divvy-tripdata.s3.amazonaws.com/index.html (from Jan-2023 till        Dec-2023) after making sure that this source is reliable, original, comprehensive, current, and cited.

### **3. Process:** Documentation of any cleaning or manipulation of data.
- **Collecting the data:**
  a.	First downloading the data (12 separate csv files as file for each month).
  b.	Unzip the files.
  c.	Create a folder on my desktop to house the files. Use appropriate file-naming conventions.
  d.	Create subfolders for the .csv file so that I have a copy of the original data. Move the downloaded files to the appropriate        subfolder.
  e.	Open RStudio to merge all the 12 csv files together in one data frame as in the below code chunk:\
  (I am using R language as the data is around 5 million records and Excel won’t handle it properly)

```{R }
library(tidyverse)
library(data.table)
df <- list.files(path = 'D:/Bike-share data/data/Bike-Share/', pattern = '*.csv') %>% 
  map_df(~fread(.))
head(df)
```
  I used the head() function to see if the data collected properly.
  

- Alternatively, we can use Python to collect the 12 csv files into one data frame as below:
    ```
    #import relevant libraries
    import pandas as pd
    import glob

    #Get a list of all CSV files in a directory
    csv_files = glob.glob('D:/Bike-share data/data/Bike-Share/*.csv')

    #Create an empty dataframe to store the combined data
    combined_df = pd.DataFrame()

    #Loop through each CSV file and append its contents to the combined dataframe
    for csv_file in csv_files:
      df = pd.read_csv(csv_file)
      combined_df = pd.concat([combined_df, df])
    print(combined_df)
    ```

-	**Manipulate the data for better understanding:**
    a. Use the colnames() function to see the column names.
        ```{R }
        colnames(df)
        ```
    b. Create a new variable called ride_length. Calculate the length of each ride by subtracting the column started_at from the column ended_at and divide it by 60 to convert to minutes, as in the below code chunk:
        ```{R }
        df <- mutate(df, ride_length=(ended_at-started_at)/60)
        ```
    c. Create a column called day_of_week, and calculate the day of the week that each ride started using the wday function, noting that 1 = Sunday and 7 = Saturday.
        ```{R }
        df <- mutate(df, day_of_week=wday(started_at))
        ```
    d. Converting the blank values to NA.
        ```{R }
        df <- df %>% mutate_at(c('ride_id','rideable_type','start_station_name','start_station_id','end_station_name','end_station_id', 'member_casual'), ~na_if(., ''))
        ```
    e. Creating another dataframe (df_cleaned) without the unwanted columns start_lat, start_lng, end_lat, end_lng, start_station_id, end_station_id because these columns won’t be needed for this analysis, and renaming rideable_type to bike_type and member_casual to customer_type.
        ```{R }
        df_cleaned <- df %>% 
        select('ride_id','rideable_type', 'started_at', 'ended_at', 'start_station_name', 'end_station_name', 'member_casual', 'ride_length', 'day_of_week') %>% 
        rename(bike_type=rideable_type, customer_type=member_casual)
        ```

- **Now we will export the data from R to MySQL database as it processes the data more quickly than RStudio.**
    a. Install the RMySQL and DBI packages and load them.
        ```{R eval=FALSE}
        install.packages('RMySQL')
        install.packages('DBI')
        library(RMySQL)
        library(DBI)
        ```

    b. Open MySQL and create an empty table on MySQL database with the same column names as following.
    
    ``` {r eval=FALSE}
      CREATE TABLE bike_share (
      ride_id		VARCHAR (45) NOT NULL,
      bike_type		VARCHAR(45),
      started_at		DATETIME,
      ended_at		DATETIME,
      start_station_name	VARCHAR(45),
      end_station_name	VARCHAR(45),
      customer_type	VARCHAR(45),
      ride_length		FLOAT,
      day_of_week		INT,
      PRIMARY KEY (ride_id));
    ```

    c. On R run the below code to connect to the MySQL database that is hosted on the local machine.
    ```{r eval=FALSE}
        con <- dbConnect(MySQL(), user="user", password="pass", 
                 dbname="database", host="localhost”)
    ```

    d. Then run the below code to view our tables that are in the database.
        ```{R eval=FALSE}
        dbListTables(con)
        ```

    e. Use the dbWriteTable() to export the R dataframe (df) to MySQL, then disconnect from the server.
        ```{R eval=FALSE}
        dbWriteTable(con, name= "bike_share", value= df, append= TRUE, temporary= FALSE, row.names = FALSE)
        dbDisconnect(con)
        ```

    f. To make sure that the data is imported successfully to MySQL database we will run the below query.
        ```{R eval=FALSE}
        SELECT 
            COUNT(*)
        FROM
            bike_share;
        ```
        ```
        mysql> 5719877 row(s) returned.
        ```
        The count result for all the records is 5,719,877 which is the same as in the R dataframe.

-	**Now it is time to run some calculations on the data to make sure it is clean and reliable.**
    a. Count the ride_length records where equal or below 0 minutes and delete those records.
        ```{R eval=FALSE}
        SELECT 
            COUNT(*)
        FROM
            bike_share
        WHERE
            ride_length <= 0;
        ```
        ```
        mysql> 528row(s) returned.
        ```

        ```
        DELETE FROM bike_share 
        WHERE
            ride_length <= 0;
        ```
        ```
        mysql> 528row(s) affected.
        ```
    b. There are records with the same station name in start_station_name and end_station_name, with ride_length less than 1 minute, which means that the customer might have changed his mind and docked the bike back to the station.
So we need to delete those records.
        ```{R eval=FALSE}
        SELECT 
            COUNT(*)
        FROM
            bike_share
        WHERE
            ride_length < 1
                AND start_station_name = end_station_name;
        ```
        ```
        mysql> COUNT(*)
               83501
        ```
        ```
        DELETE FROM bike_share
        WHERE
            ride_length < 1 AND start_station_name = end_station_name;
        ```
        ```
        mysql> 83501 row(s) affected.
        ```
    c. There are also records with null start_station_name and end_station_name and ride_length less than 1 minute, so we will delete it.
        ```{R eval=FALSE}
        SELECT 
            COUNT(*)
        FROM
            bike_share
        WHERE
            start_station_name IS NULL
                AND end_station_name IS NULL
                AND ride_length < 1;
        ```
        ```
        mysql> COUNT(*)
	             41144
        ```

        ```
        DELETE FROM bike_share 
        WHERE
            start_station_name IS NULL
            AND end_station_name IS NULL
            AND ride_length < 1;
        ```

        ```
        mysql> 41144 row(s) affected.
        ```

    d.Remove (Temp) from start_station_name and end_station_name to make it more consistent.
    ```{R eval=FALSE}
        UPDATE bike_share 
        SET 
            end_station_name = REPLACE(end_station_name, ' (Temp)', ''),
            start_station_name = REPLACE(start_station_name, ' (Temp)', '')
        WHERE
            end_station_name LIKE '% (Temp)'
                OR start_station_name LIKE '% (Temp)';
    ```

    ```{R eval=FALSE}
        mysql> 112695 row(s) affected Rows matched: 112695 Changed: 112695 Warnings: 0
    ```
    e. Search for outliers in the ride_length to avoid skewing our analysis.
We can find the outliers by calculating the interquartile range (IQR) as below:

        ```
        SELECT ride_length_quartile, MAX(ride_length) AS quartile_break
        FROM( SELECT ride_id, ride_length, NTILE(4) OVER (ORDER BY ride_length) AS ride_length_quartile
        		FROM bike_share) AS quartiles
        WHERE ride_length_quartile IN (1, 3)
        GROUP BY ride_length_quartile;
        ```
        ```
        mysql>
        ride_length_quartile	quartile_break
        1			5.7
        3			17.2167
        ```
        IQR = Q3 - Q1 = 17.2167 - 5.7\
        IQR = 11.5167\
        High outlier > Q3 + (1.5 x IQR)\
        Low outlier < Q1 - (1.5 x IQR)\
        17.2167 + (1.5 x 11.5167) = 34.49175\
        5.7 - (1.5 x 11.5167) = -11.57505\

        ```
        DELETE FROM bike_share 
        WHERE
            ride_length < - 11.57505
            OR ride_length > 34.49175;
        ```
        ```
        mysql> 412675 row(s) affected.
        ```

    f. Calculate the minimum and maximum ride_length. to make sure there are no outliers.
        ```{R eval=FALSE}
        SELECT 
            MIN(ride_length) AS min_ride_length,
            MAX(ride_length) AS max_ride_length
        FROM
            bike_share;
        ```
        ```
        mysql> min_ride_length	max_ride_length
	             0.15			34.4833
        ```

    g. Add new column Season and season_num.

        ```{R eval=FALSE}
        ALTER TABLE bike_share
        ADD COLUMN Season VARCHAR(45)
        ADD COLUMN season_num INT;
        ```
        ```
        UPDATE bike_share 
        SET 
            season = CASE
                WHEN month_number IN (12 , 1, 2) THEN 'Winter'
                WHEN month_number IN (3 , 4, 5) THEN 'Spring'
                WHEN month_number IN (6 , 7, 8) THEN 'Summer'
                WHEN month_number IN (9 , 10, 11) THEN 'Fall'
            END;
        ```

        ```
        mysql> 5160876 row(s) affected Rows matched: 5160876  Changed: 5160876  Warnings: 0
        ```

        ```
        UPDATE bike_share 
        SET 
            season_num = CASE
                WHEN season = 'Winter' THEN 1
                WHEN season = 'Spring' THEN 2
                WHEN season = 'Summer' THEN 3
                WHEN season = 'Fall' THEN 4
            END;
        ```
        ```
        mysql> 5160876 row(s) affected Rows matched: 5160876  Changed: 5160876  Warnings: 0
        ```

### **4. Analyze:** A summary of the analysis.
-	**Now it is time for some exploration.**
-	**What is the weekly average ride length for casual and members riders?**

    a. Let’s first calculate the mode for each customer_type.
    ```{R eval=FALSE}
        CREATE TEMPORARY TABLE temp 
        SELECT customer_type, day_of_week, COUNT(*) AS days
		        FROM bike_share
        		GROUP BY customer_type, day_of_week;
    ```

       ```{R eval=FALSE}
       SELECT customer_type, day_of_week
       FROM (SELECT day_of_week, COUNT(*) AS count_days
       	FROM bike_share
       	GROUP BY customer_type, day_of_week) t1,
       	(SELECT customer_type, MAX(days) AS common_day
       	FROM temp
       	GROUP BY customer_type) t2
       WHERE count_days=common_day;
       ```

       ```{R eval=FALSE}
       mysql> 14 row(s) affected Records: 14  Duplicates: 0  Warnings: 0
       mysql> customer_type	day_of_week
              casual        7
              member        5
       ```

    - **Now we know that the causal customers usually go for a ride on Sundays, while the member customers on Thursdays.**

    b. Calculate the average ride_length and group by customer_type.
      ```{R eval=FALSE}
        SELECT 
           customer_type,
           AVG(ride_length) AS avg_ride_length
        FROM
           bike_share
        GROUP BY customer_type;
      ```

       ```
       mysql> customer_type	avg_ride_length
       	       casual		    12.48
       	       member   		10.37
       ```
    c. Calculate the weekly average ride_length per customer_type.
    But first we will add week_number and month_number as we will need them in further analysis.
       ```
       ALTER TABLE bike_share
       ADD COLUMN week_number INT,
       ADD COLUMN month_number INT;
       ```
       ```
       UPDATE bike_share 
       SET 
           week_number = WEEK(started_at),
           month_number = month(started_at);
       ```
       ```
       mysql> 0 row(s) affected Records: 0  Duplicates: 0  Warnings: 0
       mysql> 5160876 row(s) affected Rows matched: 5160876  Changed: 5160876  Warnings: 0
       ```
       ```
       SELECT 
           t2.customer_type,
           AVG(avg_ride_length) AS avg_weekly_ride_length
       FROM
           (SELECT 
       	customer_type,
               week_number,
               AVG(ride_length) AS avg_ride_length
           FROM
               bike_share
           GROUP BY customer_type, week_number
           ORDER BY week_number) t2
       GROUP BY t2.customer_type
       ORDER BY avg_weekly_ride_length;
       ```
       ```
       mysql> customer_type	avg_weekly_ride_length
       	      casual		    11.625
       	      member    		10.046
       ```
    - **The result shows that the average annual and weekly ride length for Causal rides is higher than Member rides.**

-	**Is there a correlation between the seasons in Chicago and the number of rides?**
    a. First let’s create a temporary table with the daily rides count and seasons number.
        ```{R eval=FALSE}
        CREATE TEMPORARY TABLE daily_rides_count
        SELECT 
        	Season, season_num, DATE(started_at) days, COUNT(ride_id) AS ride_count
        FROM
        	bike_share
        GROUP BY season, season_num, days
        ORDER BY days;
        ```
        ```
        mysql> 365 row(s) affected Records: 365  Duplicates: 0  Warnings: 0
        ```
    b. Now we can calculate the correlation between the seasons and the number of rides.
        ```{R eval=FALSE}
        SELECT 
            (AVG(season_num * ride_count) - AVG(season_num) * AVG(ride_count)) / 
            (SQRT(AVG(season_num * season_num) - AVG(season_num) * AVG(season_num)) * 
            SQRT(AVG(ride_count * ride_count) - AVG(ride_count) * AVG(ride_count))) AS correlation
        FROM
            daily_rides_count;
        ```
        ```
        mysql> 0.588
        ```
      - The result is closer to +1 which means there is a strong correlation between the seasons and the number of rides.

-	**Is there a correlation between the seasons in Chicago and ride length?**
    a. Let’s create a temporary table with the daily rides average and seasons number.
        ```{R eval=FALSE}
        CREATE TEMPORARY TABLE avg_daily_rides
        SELECT 
        	Season, season_num, DATE(started_at) days, AVG(ride_length) AS avg_ride_length	
        FROM
        	bike_share
        GROUP BY season, season_num, days
        ORDER BY days;
        ```
        ```
        mysql> 365 row(s) affected Records: 365  Duplicates: 0  Warnings: 0
        ```
    b. Now we can calculate the correlation between the seasons and the daily average ride length.
        ```{R eval=FALSE}
        SELECT 
            (AVG(season_num * avg_ride_length) - AVG(season_num) * AVG(avg_ride_length)) / 
            (SQRT(AVG(season_num * season_num) - AVG(season_num) * AVG(season_num)) * 
            SQRT(AVG(avg_ride_length * avg_ride_length) - AVG(avg_ride_length) * AVG(avg_ride_length))) AS correlation
        FROM
            avg_daily_rides;
        ```
        ```
        mysql> 0.423
        ```
    - The above score is moderate correlation which means there is a correlation between them but not strong.

-	**What is the average number of rides on each day of the week?**
    a. Calculate the average number of rides.
        ```{R eval=FALSE}
        SELECT 
            DATE_FORMAT(daily, '%W') weekday,
            t1.day_of_week,
            ROUND(AVG(ride_num), 2) avg_ride_num
        FROM
            bike_share t1,
            (SELECT 
                DATE(started_at) daily, COUNT(ride_id) ride_num, day_of_week
            FROM
                bike_share
            GROUP BY daily , day_of_week
            ORDER BY daily) t2
        WHERE
            t1.day_of_week = t2.day_of_week
        GROUP BY weekday , t1.day_of_week
        ORDER BY t1.day_of_week;
        ```
        ```
        mysql>  weekday	    day_of_week   avg_ride_num
                Sunday      1             12147.92
                Monday      2             12754.48
                Tuesday     3             14564.71
                Wednesday   4             14878.54
                Thursday    5             15244.81
                Friday	    6             14673.48
                Saturday    7             14750.06
        ```

-	**What is the average ride length on each day of the week?**
    a. Calculate the average ride length.
        ```{R eval=FALSE}
        SELECT 
            DATE_FORMAT(started_at, '%W') weekday,
            day_of_week,
            ROUND(AVG(ride_length), 2) avg_ride_length
        FROM
            bike_share
        GROUP BY weekday, day_of_week
        ORDER BY day_of_week;
        ```
        ```
        mysql>  weekday	  day_of_week	avg_ride_length
                Sunday    1       		12.02
                Monday    2       		10.61
                Tuesday   3       		10.63
                Wednesday 4       		10.56
                Thursday  5       		10.68
                Friday    6       		10.99
                Saturday  7       		12.17
        ```
-	**Which stations do casual riders often visit?**
    a. Calculate top 10 started from stations.
        ```{R eval=FALSE}
        SELECT start_station_name, COUNT(ride_id) rides_count
        FROM bike_share
        WHERE start_station_name IS NOT NULL AND customer_type = 'casual'
        GROUP BY start_station_name
        ORDER BY rides_count DESC
        LIMIT 10;
        ```
        ```
        mysql>  start_station_name                    rides_count
                Streeter Dr & Grand Ave               30230
                DuSable Lake Shore Dr & Monroe St     19468
                DuSable Lake Shore Dr & North Blvd    15955
                Michigan Ave & Oak St                 14588
                Shedd Aquarium                        13154
                Millennium Park                       12641
                Theater on the Lake                   12298
                Wells St & Concord Ln                 11056
                Dusable Harbor                        10253
                Clark St & Elm St                     9715
        ```

b.	Calculate top 10 visited stations.
      ```{R eval=FALSE}
        SELECT end_station_name, COUNT(ride_id) rides_count
        FROM bike_share
        WHERE end_station_name IS NOT NULL AND customer_type = 'casual'
        GROUP BY end_station_name
        ORDER BY rides_count DESC
        LIMIT 10;
      ```
      ```{R eval=FALSE}
        mysql>  end_station_name                      rides_count
                Streeter Dr & Grand Ave               31244
                DuSable Lake Shore Dr & North Blvd    17808
                DuSable Lake Shore Dr & Monroe St     16690
                Michigan Ave & Oak St                 15147
                Millennium Park                       14747
                Theater on the Lake                   12254
                Shedd Aquarium                        11246
                Wells St & Concord Ln                 10954
                Clark St & Lincoln Ave                9628
                Clark St & Armitage Ave               9535
      ```

-	**What is the most used bike_type for casual riders?**
    a. Group by the bike_type with the count of rides.
        ```{R eval=FALSE}
        SELECT bike_type, COUNT(ride_id) rides_count
        FROM bike_share
        WHERE customer_type = 'casual'
        GROUP BY bike_type
        ORDER BY rides_count DESC;
        ```
        ```
        mysql>  bike_type       rides_count
                electric_bike   983407
                classic_bike    702702
                docked_bike     45048
        ```

### **5. Share: Supporting visualizations and key findings.**
- **Now with the analysis we have made so far, we can create visuals using Tableau Public to help us to identify the trends and seasonality.** 
    a. To move the data to Tableau Public we need to move it to RStudio and save it as .RData file.
        ```{R eval=FALSE}
        con <- dbConnect(MySQL(), user='user',password='pass', dbname='database',host='localhost')
        df_cleaned <- dbReadTable(con, name= 'bike_share', row.names = FALSE)
        dbDisconnect(con)
        save(df_cleaned, file = 'bike_share.RData')
        ```

    b. Then open Tableau Public and chose Statistical File, then chose bike_share.RData.

-	**Create a Dashboard with the analysis we have made in the previous steps to answer each question.**
\
\
-	**Link to the created Tableau Story:**
https://public.tableau.com/app/profile/ahmed.nasser.viz/viz/BikeShare_17087406050140/CyclisticStory

### **6. Act: The top recommendations based on my analysis.**
- **Conclusion:**
    a. Casual riders usually go for a ride on Sundays, while member riders go on Thursdays.
    b. We found that casual riders tend to ride longer distances on average than member riders. Thus, introducing different packages would likely attract casual riders to subscribe.
    c. There is a strong correlation between the number of rides and the seasons, with a noticeable increase starting from the end of the spring till the beginning of the fall, meaning that riders prefer hot weather.
    d. Ride numbers increase on weekdays, but ride durations increase on weekends, indicating that casual riders spend more time riding on weekends when they have free time to spare.
    e. Upon conducting an online search for the most visited stations, it was found that they are all located in or near parks, meaning they take the rides to visit the parks.
    f. Electric bikes are the most used by both types of riders. They are equipped with a thumb throttle and pedal assist, which makes it easier to travel further and tackle steep hills.  
\

- **Recommendations:**
    1. Launch a weekly and monthly package for riders who have a high average ride duration.
    2. Seasonal offer with a discount for new members that starts from the end of spring until summer.
    3. Create a special discount at weekends for members whose ride length is longer than 25 minutes.
    4. Design merchandise such as shirts and water bottles with the company logo in the top visited parks.
    5. Discount on parks tickets for members who arrive with Cyclistic bikes.
    6. Increase the number of electric bikes in the steep hills' areas.  
\
-	**Link to the PowerPoint presentation:**
https://github.com/a7mdNasrr/Cyclistic-bike-share/blob/main/Bike%20Share%20Presentation.pptx
\
\